On modeling for Kerr black holes: Basis learning, QNM frequencies, and spherical-spheroidal mixing coefficients

### 2019-04-02 docs/mvfits_paper/mvf.tex
Add note about Berti+Klein fit transformation [26]

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Abstract %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  %
  Models of black hole properties play an important role in the ongoing direct detection of gravitational waves from black hole binaries. One important aspect of model based gravitational wave detection, and subsequent estimation of source parameters, are the low level modeling of information related to perturbed Kerr black holes. Here, we present new phenomenological methods to model the analytically understood gravitational wave spectra (quasi-normal mode frequencies), and harmonic structure of Kerr black holes (mixing coefficients between spherical and spheroidal harmonics). In particular, we present a greedy-multivariate-polynomial (GMVP) regression method and greedy-multivariate-rational (GMVR) regression method for the automated modeling of polynomial and rational functions respectively. GMVP is used to develop a model for QNM frequencies that explicitly enforces consistency with the extremal Kerr limit. GMVR is used to develop a model for harmonic mixing coefficients for the dominant multipoles with $\ell \le 5$. The models for the mixing coefficients and $(l,m)\neq(2,2)$ QNM frequencies are the first of their kind to consider black hole spin to vary between -1 and 1, thus naturally connecting the pro and retrograde modes. We discuss the potential use of these models in current and future gravitational wave signal modeling.
	%
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In the coming years, expectations for frequent \gw{} detections of increasing \snr{} are high \cite{TheLIGOScientific:2016pea,Abbott:2016nhf,LIGOScientific:2018mvr}.

@@

Most prominently, there is ongoing interest in efficient and accurate signal models for \bbh{} \imr{} \cite{Blackman:2017dfb, London:2017bcn, Hannam:2013oca, Varma:2018mmi, Pan:2013rra, Cotesta:2018fcv}.

@@

\cite{Hannam:2013oca, London:2017bcn, Khan:2015jqa, Mehta:2017jpq,Khan:2018fmp}.

@@

\par Phenomenological models of remnant \bh{} mass and spin are used by both Phenom and \eob{} approaches either to interpolate over tables of \qnm{} frequencies or in the evaluation of phenomenological models for \qnm{} frequencies.

@@

\par In this combined context, it is clear that the modeling of \qnm{} frequencies, $\cw\lmn$, and spherical-spheroidal mixing coefficients, $\sigma\LMlmn$, are relevant for a range of \gw{} signal models (e.g. \cite{Mehta:2019wxm}).

@@

\par For both the $(l,m)\neq(2,2)$ \qnm{} frequencies and the spherical-sphoidal mixing coefficients, we present the first models which treat \qnm{s} rotating with and against the rotation of the \bh{} as being a part of a single solution parameterized by dimensionless \bh{} spin ranging from -1 to 1.

@@

In \sec{meth}, we outline the \gmvp{} and \gmvr{} algorithms.

@@

The study of continuous scalar functions often centers about their expansion in terms of monomials.

@@

This is made rigorous by the Generalised Stone-Weierstrass Theorem \cite{Hassani:2013}. Consider a continuous function $f(\vec{x}) \equiv f(x_0, x_1, ..., x_N)$ in $N+1$ variables. Then $f(\vec{x})$ can be represented as

@@

  f( \vec{x}) &= \sum_{\substack{d_i=0 \\ i \in \{0 \ldots N\}}}^{\infty} c_{d_0 d_1 \ldots d_N} x^{d_0} x^{d_1} \ldots x^{d_N}

@@

  &\approx \sum_{\substack{d_i=0 \\ i \in \{0 \ldots N\}}}^{K} c_{d_0 d_1 \ldots d_N} x^{d_0} x^{d_1} \ldots x^{d_N}

@@

From the first to second line of \eqn{eq:mvt} the equality has been replaced by an approximation as we have limited the linear representation to monomials with combined exponents no greater than $KN$. This implication follows from the Stone-Weierstrass Theorem \cite{Lang:1983} and tells us that a continuous function may be uniformly approximated by a polynomial.

@@

\par This is key to the perspective of \gmvp{}:

@@

given training data thought to be related to a continuous multivariate function, it may, particularly on small scales, be well approximated by a polynomial in an appropriate coordinate basis.

@@

\par In this setting, the uncertainty of which and how many basis terms to include makes this a problem ripe for the application of linear modeling driven by a greedy process, namely \eqn{eq:linsol1} and Alg. (\ref{alg:pgreedy}) .

@@

  The grey dashed line marks the value of $\kappa$ where $\j{}=0$. Fits of dimensionless \qnm{} decay rates (solid lines) along with select numerical values (grey markers) computed using Leaver's method \cite{Leaver85}. (Bottom) Percentage absolute residual errors for fits of dimensionless \qnm{} frequencies,, $\cw\lmn$ along with select numerical values (colored markers) computed using Leaver's method \cite{Leaver85}. For comparison, residuals using the models in Ref.~\cite{Berti:2005ys} are shown in gray and a single residual using a fit for $\cw_{220}$ from \cite{Nagar:2018zoe} is shown as an orange dashed line.}

@@

The refinement process is to terminate when a measure of model error (e.g. the $L^2$ norm $||\vec{g}^{(0)}-\vec{g}^{(n)}||$) passes a predetermined threshold. For the results presented in \sec{results} the following model representation error was used

@@

  \operatorname{Var}\left(\vec{d}\right)  = \frac{1}{J}\sum_{j=0}^{J}\left(d_j - \frac{1}{J}\sum_{u=0}^{J}d_u\right)^2

@@

We note that, in the asymptotic vicinity of $\jf=1$, the \qnm{} frequencies and decay times are known to have solutions that are asymptoticly degenerate \cite{Zimmerman:2015trm}.

@@

here, the sharp feature near $\jf=0$ results from their modeling counter and co-rotating \qnm{} as two different curves. A residual error is also shown as an orange dashed line for the $\cw_{220}$ model presented in Ref.~\cite{Nagar:2018zoe}. This is a rational model with 14 fitted parameters compared to the polynomial model with 10 fitted parameters defined in \eqn{eq:cw_fit_1}. We observe that Ref.~\cite{Nagar:2018zoe} performs better overall while performing worse for higher $\jf$.

@@

\par \gmvp{} has been applied to the modeling of \qnm{} frequencies. The resulting polynomial models have been constrained in the extremal Kerr limit, and perform well compared to similar fits in the literature. We note our comparison of $\cw_{220}$ from Ref.~\cite{Nagar:2018zoe} suggests even better fits for $(l,m,n)\neq(2,2,0)$ QNM frequencies might be achievable using a rational ansatz. We leave this extension to future modeling work.

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%% Abstract %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\begin{abstract}
  %
  Models of black hole properties play an important role in the ongoing direct detection of gravitational waves from black hole binaries. One important aspect of model based gravitational wave detection, and subsequent estimation of source parameters, are the low level modeling of information related to perturbed Kerr black holes. Here, we present new phenomenological methods to model the analytically understood gravitational wave spectra (quasi-normal mode frequencies), and harmonic structure of Kerr black holes (mixing coefficients between spherical and spheroidal harmonics). In particular, we present a greedy-multivariate-polynomial (GMVP) regression method and greedy-multivariate-rational (GMVR) regression method for the automated modeling of polynomial and rational functions respectively. GMVP is used to develop a model for QNM frequencies that explicitly enforces consistency with the extremal Kerr limit. GMVR is used to develop a model for harmonic mixing coefficients for the dominant multipoles with $\ell \le 5$. Both models are the first of their kind to consider black hole spin to vary between -1 and 1, thus naturally connecting the pro and retrograde modes. We discuss the potential use of these models in current and future gravitational wave signal modeling.
	%
\end{abstract}
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

In the coming years, expectations for frequent \gw{} detections of increasing \snr{} are high \cite{TheLIGOScientific:2016pea,Abbott:2016nhf}.

@@

Most prominently, there is ongoing interest in efficient and accurate signal models for \bbh{} \imr{} \cite{Blackman:2017dfb, London:2017bcn, Hannam:2013oca}.

@@

\cite{Hannam:2013oca, London:2017bcn, Khan:2015jqa, Schmidt:2014iyl, Mehta:2017jpq,Khan:2018fmp}.

@@

\par Both Phenom and \eob{} approaches either use phenomenological models of remnant \bh{} mass and spin to interpolate over tables of \qnm{} frequencies, or phenomenological models for \qnm{} frequencies are used directly.

@@

\par In this combined context, it is clear that the modeling of \qnm{} frequencies, $\cw\lmn$, and spherical-spheroidal mixing coefficients, $\sigma\LMlmn$, are relevant for a range of \gw{} signal models.

@@

\par For both the \qnm{} frequencies and the spherical-sphoidal mixing coefficients, we present the first models which treat \qnm{s} rotating with and against the rotation of the \bh{} as being a part of a single solution parameterized by dimensionless \bh{} spin ranging from -1 to 1.

@@

In section \sec{meth}, we outline the \gmvp{} and \gmvr{} algorithms.

@@

The study of smooth scalar functions (e.g. $f(\vec{x})$) often centers about the Taylor series expansion.

@@

In that instance, it is clear that any infinitely differentiable scalar function of many variables can be represented in terms of its derivatives by

@@

  f( \vec{x} + \vec{h} ) &= e^{ \vec{x} \cdot \vec{\nabla}' } \, f(\vec{x}\,') |_{\vec{x}\,'=\vec{h}}

@@

  &\approx \sum_{k=0}^{K} \, \frac{1}{k!} \, (\vec{x} \cdot \vec{\nabla}')^{k} \, f(\vec{x}\,') |_{\vec{x}\,'=\vec{h}}

@@

From the first to second line of \eqn{eq:mvt}, we have used the definition of the exponential function (i.e. its series expansion).

@@

In the second line, the equality has been replaced by an approximation as we have limited the linear representation to $K+1$ terms.

@@

\par This latter point is key to the perspective of \gmvp{}:
given training data thought to be related to a smooth multivariate function, it may, particularly on small scales, be well approximated by a truncated series expansion in an appropriate coordinate basis.

@@

\par \eqn{eq:mvt}'s truncated expansion happens to be a polynomial in at most $N$ variables.
In this setting, the uncertainty of which and how many basis terms to include makes this a problem ripe for the application of linear modeling driven by a greedy process, namely \eqn{eq:linsol1} and Alg. (\ref{alg:pgreedy}) .

@@

  The grey dashed line marks the value of $\kappa$ where $\j{}=0$. Fits of dimensionless \qnm{} decay rates (solid lines) along with select numerical values (grey markers) computed using Leaver's method \cite{Leaver85}. (Bottom) Percentage absolute residual errors for fits of dimensionless \qnm{} frequencies,, $\cw\lmn$ along with select numerical values (colored markers) computed using Leaver's method \cite{Leaver85}. For comparison, residuals using the models in Ref.~\cite{Berti:2005ys} are shown in gray. }

@@

The refinement process is to terminate when a measure of model error (e.g. the $L^2$ norm $||\vec{g}^{(0)}-\vec{g}^{(n)}||$) passes a predetermined threshold. For \cite{lionel_london_2018_1402516} the following model representation error was used

@@

  \operatorname{Var}\left(\vec{a}\right)  = \frac{1}{J}\sum_{j=0}^{J}\left(d_j - \frac{1}{J}\sum_{u=0}^{J}d_u\right)^2

@@

We note that, in the asymptotic vicinity of $\jf=1$, the \qnm{} frequencies and decay times are known to have solutions that are asymptoticly degenerate \cite{Zimmerman:2015rua}.

@@

here, the sharp feature near $\jf=0$ results from their modeling counter and co-rotating \qnm{} as two different curves.

@@

\par \gmvp{} has been applied to the modeling of \qnm{} frequencies. The resulting models have been constrained in the extremal Kerr limit, and perform well compared to similar fits in the literature.
